---
title: "NOAA Storm Database Analysis"
author: "André Marinho"
date: "25/07/2020"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Synopsis
COLOCAR NA SINOPSE:

Which types of events are most harmful with respect to population health, and which types of events have the greatest economic consequences across the United States


## 2. Data Processing
Steps to load dataset into R (it did not any preprocessing outside this document):

1. Download dataset;
2. Unzip the csv file representing dataset;
3. Read the csv file;
4. Note the first lines of csv file;
5. Get the names of required health and economic analysis data;
6. Analysis of the data types;
7. Preprocessing: treat PROPDMGEXP values (PROPDMG = PROPDMG * PROPDMGEXP_preprocessed)
8. Preprocessing: treat CROPDMGEXP values (CROPDMG = CROPDMG * CROPDMGEXP_preprocessed)

Steps 3 and 4
```{r cache = TRUE}
data <- read.csv("../repdata_data_StormData.csv", sep=",", header=TRUE)
head(data)
```
Steps 5 and 6
```{r cache = TRUE}
names(data)
data_required <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
data <- data[data_required]
head(data)
```

```{r}
#unique(data$EVTYPE)
#unique(data$FATALITIES)
#unique(data$INJURIES)
#unique(data$PROPDMG)
unique(data$PROPDMGEXP)
```

Step 7 (records like "+", "-", "?" were treated like 0)
```{r message=FALSE, warning=FALSE}
library(dplyr)
data <- data %>%
        mutate(PROPDMG = PROPDMG * case_when(
                PROPDMGEXP == "H" ~ 1e+02,
                PROPDMGEXP == "h" ~ 1e+02,
                PROPDMGEXP == "K" ~ 1e+03,
                PROPDMGEXP == "M" ~ 1e+06,
                PROPDMGEXP == "m" ~ 1e+06,
                PROPDMGEXP == "B" ~ 1e+09,
                PROPDMGEXP == "0" ~ 1,
                PROPDMGEXP == "1" ~ 10,
                PROPDMGEXP == "2" ~ 1e+02,
                PROPDMGEXP == "3" ~ 1e+03,
                PROPDMGEXP == "4" ~ 1e+04,
                PROPDMGEXP == "5" ~ 1e+05,
                PROPDMGEXP == "6" ~ 1e+06,
                PROPDMGEXP == "7" ~ 1e+07,
                PROPDMGEXP == "8" ~ 1e+08,
                PROPDMGEXP == "" ~ 1,
                PROPDMGEXP == "+" ~ 0,
                PROPDMGEXP == "-" ~ 0,
                PROPDMGEXP == "?" ~ 0
        ), PROPDMGEXP=NULL
        )
```

Step 6, again (other columns)
```{r}
#unique(data$CROPDMG)
unique(data$CROPDMGEXP)
```

Step 8 (records like "+", "-", "?" were treated like 0)
```{r}
data <- data %>%
        mutate(CROPDMG = CROPDMG * case_when(
                CROPDMGEXP == "K" ~ 1e+03,
                CROPDMGEXP == "k" ~ 1e+03,
                CROPDMGEXP == "M" ~ 1e+06,
                CROPDMGEXP == "m" ~ 1e+06,
                CROPDMGEXP == "B" ~ 1e+09,
                CROPDMGEXP == 0 ~ 1,
                CROPDMGEXP == 2 ~ 1e+02,
                CROPDMGEXP == "" ~ 1,
                CROPDMGEXP == "?" ~ 0
        ), CROPDMGEXP=NULL
        )
```

## 3. Data Analysis
Calculate the totals of each incident by event type

1. Population health events: fatalities and injuries top 10;
2. Plot top 10 health issues
3. Economic problem events: property and crop damages top 10;
4. Plot top 10 economic issues
```{r}
head(data)

# 3.1. Population health events
health_events_by_fatality <- aggregate(FATALITIES ~ EVTYPE, data, FUN=sum)

health_events_by_injury <- aggregate(INJURIES ~ EVTYPE, data, FUN=sum)

fatalities_top10 <- health_events_by_fatality[with(health_events_by_fatality, order(-FATALITIES)),]

injuries_top10 <- health_events_by_injury[with(health_events_by_injury, order(-INJURIES)),]

population_issues_top10 <- cbind(
        head(fatalities_top10, 10),
        head(injuries_top10, 10))

print(population_issues_top10)
```

```{r message=FALSE, warning=FALSE}
# 3.2. Plot top 10 health issues
library(ggplot2)

fat_top10_plot <- head(population_issues_top10[1:10,1:2], 10)
inj_top10_plot <- head(population_issues_top10[1:10,3:4], 10)

par(mfrow=c(1,2),mar=c(10,4,3,2))
barplot(fat_top10_plot$FATALITIES,
        names.arg=fat_top10_plot$EVTYPE,
        las=3,
        ylab="# OF FATALITIES",
        col="red",
        main="TOP 10 FATALITIES")
barplot(inj_top10_plot$INJURIES,
        names.arg=inj_top10_plot$EVTYPE,
        las=3,
        ylab="# OF INJURIES",
        col="red",
        main="TOP 10 INJURIES")
```


```{r}
# 3.3. Economic Problem Events
economic_problems_by_property <- aggregate(PROPDMG ~ EVTYPE, data, FUN=sum)

economic_problems_by_crop <- aggregate(CROPDMG ~ EVTYPE, data, FUN=sum)

property_top10 <- economic_problems_by_property[with(economic_problems_by_property, order(-PROPDMG)),]

crop_top10 <- economic_problems_by_crop[with(economic_problems_by_crop, order(-CROPDMG)),]


economic_issues_top10 <- cbind(
        head(property_top10, 10),
        head(crop_top10, 10))

print(economic_issues_top10)
```

```{r}
# 3.4. Plot top 10 economic issues
prop_top10_plot <- head(economic_issues_top10[1:10,1:2], 10)
crop_top10_plot <- head(economic_issues_top10[1:10,3:4], 10)

par(mfrow=c(1,2),mar=c(12,4,3,2))
barplot(prop_top10_plot$PROPDMG/(1e+09),
        names.arg=prop_top10_plot$EVTYPE,
        las=3,
        ylab="(IN BILLIONS)",
        col="yellow",
        main="TOP 10 PROPERTY DAMAGES")
barplot(crop_top10_plot$CROPDMG/(1e+09),
        names.arg=crop_top10_plot$EVTYPE,
        las=3,
        ylab="(IN BILLIONS)",
        col="yellow",
        main="TOP 10 CROP DAMAGES")
```