---
title: "NOAA Storm Database: Adverse Weather Impacts Analysis for US population and economy "
author: "André Marinho"
date: "25/07/2020"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Synopsis
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The main objective of this project is to answer the following questions using the NOAA storm database:

1. Which types of events are most harmful with respect to population health?
2. Which types of events have the greatest economic consequences across the United States?

It was selected the following incidents to analysis and to answer these questions: fatalities and injuries (in the case of population health), and property and crop damage (in the case of economic consequences).

The analysis discovered that tornados are the most harmful event across the United States with respect to population health. Floods leads to the greatest economic consequences to properties, and droughts leads to the greatest economic consequences to crops.

## 2. Data Processing
Steps to load dataset into R (it did not any preprocessing outside this document):

1. Download dataset;
2. Unzip the csv file representing dataset;
3. Read the csv file;
4. Note the first lines of csv file;
5. Get the names of required health and economic analysis data;
6. Analysis of the data types;
7. Preprocessing: treat PROPDMGEXP values (PROPDMG = PROPDMG * PROPDMGEXP_preprocessed);
8. Preprocessing: treat CROPDMGEXP values (CROPDMG = CROPDMG * CROPDMGEXP_preprocessed).

Steps 3 and 4
```{r cache = TRUE}
data <- read.csv("../repdata_data_StormData.csv", sep=",", header=TRUE)
head(data)
```
Steps 5 and 6
```{r cache = TRUE}
names(data)
data_required <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
data <- data[data_required]
head(data)
```

```{r}
#unique(data$EVTYPE)
#unique(data$FATALITIES)
#unique(data$INJURIES)
#unique(data$PROPDMG)
unique(data$PROPDMGEXP)
```

Step 7 (records like "+", "-", "?" were treated as 0)
```{r message=TRUE, warning=TRUE}
library(dplyr)
data <- data %>%
        mutate(PROPDMG = PROPDMG * case_when(
                PROPDMGEXP == "H" ~ 1e+02,
                PROPDMGEXP == "h" ~ 1e+02,
                PROPDMGEXP == "K" ~ 1e+03,
                PROPDMGEXP == "M" ~ 1e+06,
                PROPDMGEXP == "m" ~ 1e+06,
                PROPDMGEXP == "B" ~ 1e+09,
                PROPDMGEXP == "0" ~ 1,
                PROPDMGEXP == "1" ~ 10,
                PROPDMGEXP == "2" ~ 1e+02,
                PROPDMGEXP == "3" ~ 1e+03,
                PROPDMGEXP == "4" ~ 1e+04,
                PROPDMGEXP == "5" ~ 1e+05,
                PROPDMGEXP == "6" ~ 1e+06,
                PROPDMGEXP == "7" ~ 1e+07,
                PROPDMGEXP == "8" ~ 1e+08,
                PROPDMGEXP == "" ~ 1,
                PROPDMGEXP == "+" ~ 0,
                PROPDMGEXP == "-" ~ 0,
                PROPDMGEXP == "?" ~ 0
        ), PROPDMGEXP=NULL
        )
```

Step 6, again (other columns)
```{r}
#unique(data$CROPDMG)
unique(data$CROPDMGEXP)
```

Step 8 (records like "+", "-", "?" were treated as 0)
```{r}
data <- data %>%
        mutate(CROPDMG = CROPDMG * case_when(
                CROPDMGEXP == "K" ~ 1e+03,
                CROPDMGEXP == "k" ~ 1e+03,
                CROPDMGEXP == "M" ~ 1e+06,
                CROPDMGEXP == "m" ~ 1e+06,
                CROPDMGEXP == "B" ~ 1e+09,
                CROPDMGEXP == 0 ~ 1,
                CROPDMGEXP == 2 ~ 1e+02,
                CROPDMGEXP == "" ~ 1,
                CROPDMGEXP == "?" ~ 0
        ), CROPDMGEXP=NULL
        )
```

## 3. Data Analysis
Calculate the totals of each incident by event type.

1. Population health events: fatalities and injuries top 10;
2. Plot top 10 health issues;
3. Economic problem events: property and crop damages top 10;
4. Plot top 10 economic issues.
```{r}
head(data)

# 3.1. Population health events
health_events_by_fatality <- aggregate(FATALITIES ~ EVTYPE, data, FUN=sum)

health_events_by_injury <- aggregate(INJURIES ~ EVTYPE, data, FUN=sum)

fatalities_top10 <- health_events_by_fatality[with(health_events_by_fatality, order(-FATALITIES)),]

injuries_top10 <- health_events_by_injury[with(health_events_by_injury, order(-INJURIES)),]

population_issues_top10 <- cbind(
        head(fatalities_top10, 10),
        head(injuries_top10, 10))

print(population_issues_top10)
```

```{r message=TRUE, warning=TRUE}
# 3.2. Plot top 10 health issues
library(ggplot2)

fat_top10_plot <- head(population_issues_top10[1:10,1:2], 10)
inj_top10_plot <- head(population_issues_top10[1:10,3:4], 10)

par(mfrow=c(1,2),mar=c(12,4,3,2))
barplot(fat_top10_plot$FATALITIES,
        names.arg=fat_top10_plot$EVTYPE,
        las=3,
        ylab="# OF FATALITIES",
        col="red",
        main="TOP 10 FATALITIES")
barplot(inj_top10_plot$INJURIES,
        names.arg=inj_top10_plot$EVTYPE,
        las=3,
        ylab="# OF INJURIES",
        col="red",
        main="TOP 10 INJURIES")
```

### Tornado is the most harmful event across the United States with respect to population health. It represents 5.633 fatalities cases, and 91.3456 injuries cases.

```{r}
# 3.3. Economic Problem Events
economic_problems_by_property <- aggregate(PROPDMG ~ EVTYPE, data, FUN=sum)

economic_problems_by_crop <- aggregate(CROPDMG ~ EVTYPE, data, FUN=sum)

property_top10 <- economic_problems_by_property[with(economic_problems_by_property, order(-PROPDMG)),]

crop_top10 <- economic_problems_by_crop[with(economic_problems_by_crop, order(-CROPDMG)),]


economic_issues_top10 <- cbind(
        head(property_top10, 10),
        head(crop_top10, 10))

print(economic_issues_top10)
```

```{r}
# 3.4. Plot top 10 economic issues
prop_top10_plot <- head(economic_issues_top10[1:10,1:2], 10)
crop_top10_plot <- head(economic_issues_top10[1:10,3:4], 10)

par(mfrow=c(1,2),mar=c(12,4,3,2))
barplot(prop_top10_plot$PROPDMG/(1e+09),
        names.arg=prop_top10_plot$EVTYPE,
        las=3,
        ylab="(IN BILLIONS)",
        col="yellow",
        main="TOP 10 PROPERTY DAMAGES")
barplot(crop_top10_plot$CROPDMG/(1e+09),
        names.arg=crop_top10_plot$EVTYPE,
        las=3,
        ylab="(IN BILLIONS)",
        col="yellow",
        main="TOP 10 CROP DAMAGES")
```

### Flood events has the greatest economic consequences to property, representing US\$144.657.709.807 of loss. Drought events has the greatest economic consequences to crop, representing US\$13.972.566.000 of loss.

## 4. Results
- **Question 1:** Across the United States, which types of events are most harmful with respect to population health?

**Answer: Tornado** is the most harmful event across the United States with respect to population health. It represents 5.633 fatalities cases, and 91.3456 injuries cases.

Specifically with respect **fatalities**, tornado event is followed by these most harmful events:

2. Excessive heat
3. Flash flood
4. Heat
5. Lightning
6. TSTM wind
7. Flood
8. RIP current
9. High wind
10. Avalanche

Specifically with respect **injuries**, tornado event is followed by these most harmful events:

2. TSTM wind
3. Flood
4. Excessive heat
5. Lightning
6. Heat
7. Ice storm
8. Flash Flood
9. Thunderstorm wind
10. Hail

- **Question 2:** Across the United States, which types of events have the greatest economic consequences?

**Answer: Flood** events has the greatest economic consequences to property, representing US\$144.657.709.807 of loss. **Drought** events has the greatest economic consequences to crop, representing US\$13.972.566.000 of loss.

Specifically with respect **properties**, flood event is followed by these most harmful events:

2. Hurricane/Typhoon
3. Tornado
4. Storm Surge
5. Flash flood
6. Hail
7. Hurricane
8. Tropical storm
9. Winter storm
10. High wind

Specifically with respect **crops**, drought event is followed by these most harmful events:

2. Flood
3. River flood
4. Ice storm
5. Hail
6. Hurricane
7. Hurricane/Typhoon
8. Flash flood
9. Extreme cold
10. Frost/Freeze
